---
title: "Automatic or Manual? A Fuel Consuption Analysis"
author: "Oszkar Jozsa"
output: pdf_document
geometry: margin=1.2cm
fontsize: 10pt
---

# Summary

In this paper, I will show that, in general, cars with automatic transmission have worse fuel economy 
compared to the manual transmission ones. However, my analysis also will show that the worse mileage 
performance is determined by other factors also, not only the type of transmission the car has.

After a quick analysis in the first section, I will construct a linaer prediction model 
that predicts the fuel consumption based on those factors.

# Analysis

## Data processing

First of all, let's load the data and have a quick look at it.

```{r}
library(ggplot2)
data(mtcars)
str(mtcars)
```

We have 32 observations. Our two important variables are `mpg` and `am`. First one is a numeric value 
of the fuel consumption, second one is the gear shift type. [This site](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) 
tells us that 0 = automatic and 1 = manual.

Lat's convert that numerical value into a factor:

```{r}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("Automatic", "Manual")
```

## Exploratory analysis

First of all, check if there is any difference between automatic and manual transmission with respect to MPG at all. 
A boxplot can clearly visualize this for us.

```{r}
boxplot(mpg~am, data=mtcars,
        xlab = "Transmission type",
        ylab = "Miles / Gallon",
        main = "Fuel consuption by transmission type")
```

Vehicles with automatic transmission seem to have lower MPG, or in other words, they have worse fuel consumption.

Another validation of this initial assumption is to break down these categories further. 
It's a good intuition that engine displacement and weight can effect fuel consuption, we should check whether these parameters are 
similar in case of automatic and manual transmissions.

```{r}
summary(mtcars[mtcars$am == "Automatic",]$disp)
summary(mtcars[mtcars$am == "Manual",]$disp)
summary(mtcars[mtcars$am == "Automatic",]$wt)
summary(mtcars[mtcars$am == "Manual",]$wt)
```

It shows that automatic shift cars tend to have larger engine displacement and larger weight, too. 
This suggests that the larger fuel consumption of automatic shift cars might not caused by the shift type itself.
Of course, other factors can effect fuel consuption but most likely we have to include these variables 
also on our final predicition model. Appendix 1 and Appendix 2 show the displacement and weight 
versus MPG per transmission type diagrams.

One final thing we can check to make sure the data is valid: it doesn't seem to have outliers and 
it seems to be roughly normally distributed. So this is not skewing our data either. 
See Appendix 3 for the distribution analysis.

## Regression model

### Simple regression

The simplest model is to use a single variable regression, using the transmission type, `am`.

```{r}
singlelm <- lm(mpg~am, data=mtcars)
summary(singlelm)
```

Turns out (unsurprisingly) that the single variable of the transmission type is not that great of a predictor 
of fuel consumption. We can see that using an automatic transmission, our fuel consumption tends to be 7 MPG higher 
compared to manual transmission. However, the R^2 value tells us that this model only explains 36% of the variance. 
Also, the squared error seems to be high, however, p-value is under 0.05.

### Multivariare regression

I will use a 2-way stepwise selection to build the model. Details of stepwise regression: 
[WikiPedia](http://en.wikipedia.org/wiki/Stepwise_regression), 
[stack exchange](http://stats.stackexchange.com/questions/69452/how-does-stepwise-regression-work), 
[Coastal Carolina University tutorial](http://ww2.coastal.edu/kingw/statistics/R-tutorials/multregr.html).

```{r}
stepmodel = step(lm(data=mtcars, mpg~.), trace=0, steps=10000)
summary(stepmodel)
```

So it seems that the transmission type is only the third best predictor variable in the final model. 
Both weight and acceleration are much better predictors 
(the latter one strongly relates to displacement size and cylinder number).

```{r}
multiplelm <- lm(mpg ~ wt + qsec + am, data=mtcars)
summary(multiplelm)
```

In the final model, the R^2 value is up to 85% and the p-value decreased well belo 0.05.

# Results

We have found a multiple variable regression model that covers 85% of the variable in the fuel consumption, 
based on the `mtcars` dataset.

The three variables in the model:

- wt - the weight of the car
- qsec - acceleration time (0.25 mile)
- am - transmission type (automatic/manual)

The performance of this model compared to a single variable model:

```{r}
rmse_single <- sqrt(sum(singlelm$residuals ^ 2) / nrow(mtcars))
rsquared_single <- summary(singlelm)$r.squared
rmse_multiple <- sqrt(sum(multiplelm$residuals ^ 2) / nrow(mtcars))
rsquared_multiple <- summary(multiplelm)$r.squared
perf <- matrix(c(rmse_single, rmse_multiple, rsquared_single, rsquared_multiple), nrow=2, ncol=2)
rownames(perf) <- c("single variable", "multiple variable")
colnames(perf) <- c("RMSE", "R^2")
perf
```

# Appendix

## Appendix 1 - Engine displacement

```{r}
dispplot <- qplot(x=disp, y=mpg, data=mtcars, colour=am, 
                  main="MPG per Engine displacement and Transmission type", 
                  xlab="Engine displacement", 
                  ylab="Miles / Gallon")
dispplot + labs(colour="Transmission type")
```

It can be seen that even though automatic shift cars have bigger engines and bigger fuel consuption also.

## Appendix 2 - Vehicle weight

```{r}
wtplot <- qplot(x=wt, y=mpg, data=mtcars, colour=am, 
                  main="MPG per Vehicle weight and Transmission type", 
                  xlab="Weight (lb/1000)", 
                  ylab="Miles / Gallon")
wtplot + labs(colour="Transmission type")
```

It can be seen that even though automatic shift cars are heavier and have bigger fuel consumption.

## Appendix 3 - MPG distribution

```{r}
mpg <- mtcars$mpg
h <- hist(mpg, breaks=11, xlab="Miles / Gallon",  main="Histogram of Miles / Gallon")
abline(v = mean(mpg), col="red", lwd=2)
abline(v = median(mpg), col="green", lwd=2)
xfit <- seq(min(mpg), max(mpg), length=40)
yfit <- dnorm(xfit, mean=mean(mpg), sd=sd(mpg))
yfit <- yfit * diff(h$mids[1:2]) * length(mpg)
lines(xfit, yfit, col="blue", lwd=2)
legend(x="topright", legend=c("Mean","Median","Normal curve"), col=c("red", "green","blue"), bty="n", lwd=1)
```