---
output:
  pdf_document: default
geometry: margin=1.1cm
fontsize: 10pt
---

#Statistical Inference Course Project - Part 1

-------------------------------------------------------------
Author       Date                   GitHub
------------ ---------------------- -------------------------
Oszkar Jozsa February, 2015         [[link][github]]
-------------------------------------------------------------

##Overview

###Description

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials.  You should

- Show the sample mean and compare it to the theoretical mean of the distribution.
- Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
- Show that the distribution is approximately normal.

###Solution

I will save a simple data set of 1000 simulations into a matrix and compare it to a theoretical normal distribution 
using some plots (base and ggplot) and simple calculations.

##Solution

###Simulations

First, we set up the parameters `number`, `lambda` and `n`. Then we'll set the seed for reproducibility.
(also, ggplot is loaded here as we'll need it later)

```{r}
library(ggplot2)

number <- 1000
lambda <- 0.2
n <- 40

set.seed(543)
```

Next, the simulation is done with the parameters above and saved into `simulation` (1000-by-40 matrix).
We also calculate the rowmeans here.

```{r}
simulation <- matrix(rexp(number * n, rate=lambda), number, n)
means <- rowMeans(simulation)
```

We create a simple histogram to visualize the result of the simulation.
Also, a red line is added for displaying the mean of the simulation.

```{r fig.height=5, fig.width=10}
hist(means, xlab = "Rowmeans of simulated matrix")
abline(v = mean(means), col = "red", lwd = 2)
legend(x = "topright", legend = c("Mean of simulation"), col = c("red"), bty = "n", lwd = 2)

```

###Sample Mean versus Theoretical Mean

```{r echo=TRUE}
mean <- mean(means)
theoretical_mean <- 1 / lambda
mean
theoretical_mean
```

Simulated (measured) mean is very close to the theoreticcal mean. 
We could get even closer by increasing `number` (more simulation).

###Sample Variance versus Theoretical Variance

```{r}
variance <- var(means)
theoretical_variance <- (1 / lambda) ^ 2 / n
variance
theoretical_variance
```

Variance differs a bit more than the mean did, but still close to the theoretical one.

###Distribution

The histogram below shows how close the distribution is to normal. Blue line is a density estimation of 
our measured data. Red line is a true normal distribution.

```{r fig.height=5, fig.width=10, message=F}
ggplot(data.frame(means), aes(x = means)) + 
      geom_histogram(aes(y=..density..)) + geom_density(colour = "blue", size = 1) +
      stat_function(fun = dnorm, args = list(mean = theoretical_mean, sd = theoretical_variance), color = "red", size = 1) +
      labs(y = "Frequency") + labs(title = "Distribution: simulated vs. normal")
```

Lastly, a QQ plot to compare measured and theoretical distribution (see [this][wiki] article). 
Lines are almost exacly lined up which shows us that the measured is very close to the theoretical one. 
A 45 degree `abline` represents the theoretical QQ line.

```{r fig.height=4.5, fig.width=10}
qqnorm(scale(means)); qqline(scale(means), col = "blue"); abline(0, 1, col = "red")
legend(x = "topleft", legend = c("Simulation", "Theoretical"), col = c("blue", "red"), bty = "n", lwd = 2)
```

[wiki]:http://en.wikipedia.org/wiki/Q-Q_plot
[github]:https://github.com/Oszkar/