---
output:
  pdf_document: default
geometry: margin=1.1cm
fontsize: 10pt
---

#Statistical Inference Course Project - Part 2

-------------------------------------------------------------
Author       Date                   GitHub
------------ ---------------------- -------------------------
Oszkar Jozsa February, 2015         [[link][github]]
-------------------------------------------------------------

##Overview

###Description

Now in the second portion of the class, we're going to analyze the ToothGrowth data in the R datasets package. 

- Load the ToothGrowth data and perform some basic exploratory data analyses 
- Provide a basic summary of the data.
- Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
- State your conclusions and the assumptions needed for your conclusions. 

##Solution

###Exploratory data analysis

We can find a descriptive info about the "The Effect of Vitamin C on Tooth Growth in Guinea Pigs" dataset [here][desc]. It says: 
"The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid)."

Let's look at the content of `ToothGrowth`!

```{r}
library(ggplot2)
library(datasets)
data(ToothGrowth)
head(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
```

So we have 2 different supplements, `VC` and `OJ` and three different doses, `0.5`, `1.0`, `2.0` (mg). 
These two (supplement and dosis) have effect on the length, saved in the `len` column. 
We can visualize the relations between these using a boxplot.

```{r}
boxplot(len ~ supp * dose, data = ToothGrowth, ylab="Tooth length", main="ToothGrowth data supplements * dosis")
```

Now, setting aside the dosage, we'll look at the effectiveness of the two supplements.

```{r}
OJ = c(mean(ToothGrowth[ToothGrowth$supp == "VC",]$len), sd(ToothGrowth[ToothGrowth$supp == "VC",]$len)) 
VC = c(mean(ToothGrowth[ToothGrowth$supp == "OJ",]$len), sd(ToothGrowth[ToothGrowth$supp == "OJ",]$len)) 
data.frame(OJ, VC, row.names = c("mean", "variance"))
```

Based on both the full box plot and on the means-variances of the two different supplements, 
we can conclude that the Orange Juice (`OJ`) supplement is better.

###Hypothesis test

T distribution will be used for hypothesis testing.

```{r}
t.test(len ~ supp, data = ToothGrowth)
```

We won't reject the null hypothesis as the P-value is above 0.05 and the confidence interval contains 0. 
This means that the effect of the two supplements fo differ.

Now test the doses (as dosage pairs):

```{r}
t.test(len ~ dose, data = subset (ToothGrowth, dose %in% c(0.5, 1.0)))
t.test(len ~ dose, data = subset (ToothGrowth, dose %in% c(0.5, 2.0)))
t.test(len ~ dose, data = subset (ToothGrowth, dose %in% c(1,0, 2.0)))
```

###Conclusions

The quick research of the `ToothGrowth` data showed us that both the dosage and the supplement type plays a role in 
tooth growth in gueina pigs. Orange juice or ascorbic acid delivery method proves to be more effective than Vitamin C. Also, bigger 
dosage helps growth (on the examined [0.5..2.0] interval).

[desc]:https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html
[github]:https://github.com/Oszkar/